Sequence learning and recall in recurrent neural networks with memory traces

### 2019-01-18 ESSAN/article.tex
last batch updated

\documentclass{esannV2}
\usepackage[dvips]{graphicx}
\usepackage[latin1]{inputenc}
\usepackage{amssymb,amsmath,array}
\graphicspath{{images/}}

@@

\begin{document}
%style file for ESANN manuscripts
\title{LaTeX style file for ESANN manuscripts}

%***********************************************************************
% AUTHORS INFORMATION AREA
%***********************************************************************
\author{Ram\'on H. Mart\'inez-Mayorquin $^1$, Anders Lansner and $^2$ Pawel Herman $^3$
%
% Optional short acknowledgment: remove next line if non-needed
% \thanks{This is an optional funding source acknowledgement.}
%
% DO NOT MODIFY THE FOLLOWING '\vspace' ARGUMENT
\vspace{.3cm}\\
%
% Addresses and institutions (remove "1- " in case of a single institution)
1- KTH - Dept of First Author \\
Address of First Author's school - Country of First Author's
school
%
% Remove the next three lines in case of a single institution
\vspace{.1cm}\\
2- KTH - Dept of Second Author \\
Address of Second Author's school - Country of Second Author's school 
%
\vspace{.1cm}\\
3 School of Second Author - Dept of Second Author \\
Address of Second Author's school - Country of Second Author's school\\ 
}

@@

\begin{abstract}
Type your 100 words abstract here. Please do not modify the style
of the paper. In particular, keep the text offsets to zero when
possible (see above in this `ESANNV2.tex' sample file). You may
\emph{slightly} modify it when necessary, but strictly respecting
the margin requirements is mandatory (see the instructions to
authors for more details).
\end{abstract}

\section{Introduction}

This is a sample file. Please use this file to correctly typeset a
submission to the ESANN conference. The associated pdf file will
help you to have an idea of what your paper should look like.

\section{Methods}
The neural network is composed of firing rate units with the following properties. First the individual units receive current from all the other units, that is, the system is a recurrently connected neural network with all-to-all connectivity. All the neurons send current to each other and the accumulated current is then  thresholded $\theta$ by the logistic function $\Phi(x) = \frac{1}{1 + \exp(-Gx)}$. This whole process is condensed into equation \ref{eq1}. The current then is integrated with a time constant $\tau_m$, the integration time of the unit, to determine the activation of the unit. In order to account for the temporal effects we introduce the idea of z-filters that keep information about the activity during a time period $\tau_z$. We can see in equation \ref{eq2} that the filter z follows the activity of the unit at any point in time.

@@

This system allows for the units to keep influencing the other units through $W$ even when they are not active anymore. We will see how this is fundamental for the sequence processing capabilities of our system.

The connectivity of the system is as depicted in figure \ref{Fig:diagram}. Note that that from each unit emanate two excitatory connections, one that self-reinforces the unit activity $Exc_{self}$ and another that excites the following neuron (contributing to the sequential structure of the activity) $Exc_T$. At the same time every unit inhibits all the units except the next one with $Inh$. Of special relevance are the backwards inhibitory projections that suppress the unit once the next unit in the sequence is activated

\begin{figure}[h!]
\centering
\includegraphics[scale=1.5]{diagram.eps}
\caption{In this diagram we can appreciate the overall connectivity of the system. Note that the inhibitory connections are only fully depicted for the first unit for the sake of clarity.}\label{Fig:diagram}
\end{figure}

@@

In order to recall a particular sequence we cue the first element. This means that we artificially activate the first unit (clamping) of the sequence for a short time. After this initial cue, if the weights are set right, the neural network continues evolving on its own and a particular sequence is recalled. In figure \ref{Fig:recall} we can appreciate a successful recall of a sequence.

\begin{figure}[h!]
\centering
\includegraphics[scale=0.20]{recall.eps}
\caption{A succesful recall. a) The activity of the units as it evolves in time. b) The currents that each unit receives as the sequence progresses. c) the z-filters activations for each unit. d) The connectivity matrix $W$. The weights of the matrix can be divided on the excitatory connections: the ones that induce transitions $Exc_{T}$ and the ones that stabilize the unit $Exc_{self}$ and the inhibitory connections $Inh$.}\label{Fig:recall}
\end{figure}

%It is important to note that the weights of the connectivity matrix determine the particular behavior that the network will develop. In the diagonal of the connectivity matrix we have the weights of the excitation connections that keep the unit active once it already became activited $Exc_{self}$. That is, they sustain a positive feed-back loop that keeps the unit active. Below the diagonal, on the other hand, we have the weights of the connections that make the sequence transition from one state to the next $Exc_{T}$. Finally, the rest of the matrix consists of the weights for the inhibitory connections. Of special significance are the backward inhibitory projections above the diagonal, their role is to suppress the last pattern of the sequence once the next unit is activated.

It is now instructive to detail at the level of the dynamics how the sequential activity is propagated. After the first unit $x_1$ has been activated for long enough due to clamping the filter $z_1$ will start catching up with the activity by increasing its value. When this happens the unit corresponding to $x_2$ will receive a growing current with value $Exc_{T} \times z_1$. As this process continues there will a point in time (exactly at $Exc_{T}(1 - e^{-\frac{t}{\tau_z}}) - \theta = 0$) when $x_2$ will receive enough current to become activated. This, will again will prompt the filter $z_2$ to start catching up initializing the process described again. On the other hand as soon as the second filter starts growing $z_2$ the first unit will start receiving an inhibitory current $Inh \times z_2$ due to the backward inhibitory projection, this will in turn suppress the unit $x_1$ allowing the sequence to propagate. Finally, once the unit $x_1$ is finally suppressed the excitatory current mediated by $Exc_T$ will be diminished which in most cases would also suppress the activity of $x_2$ if were not for the self-sustaining current $Exc_{self} \times z_2$. In short, the term $Exc_T$ ensures the transition of the activity from one unit of the sequence to the next whereas the competition between $Exc_{self}$ and $Inh$ first lock the pattern unit and then finally suppress it allowing the sequence to evolve towards its end without hindrance.

With the knowledge above in mind we can envision the following  failure scenarios for the sequential dynamics. First, if the magnitude of $Exc_{self}$ the activity will not be big enough to overcome the threshold and the sequence will not propagate. Second, if $Exc_{Self}$ is enough to keep the unit active but $Exc_{T}$ is not big enough the next unit in the sequence will not be activated and the sequence will not propagate. If both $Exc_{self}$ is big enough to sustain the activity and $Exc_{T}$ is big enough to propagate it but the inhibitory connection $Inh$ is not enough to suppress the units the result will be that all the units in the sequence will be activated at the end.

@@

A common problem in sequence learning is to scale the time duration of the sequence according to the needs of the required context. In terms of a neural network it means that once the dynamics of sequence recalling are stored in the shape of the weights we can control the \textit{recall time} of the sequence with the internal parameters of the network where the recall time is the amount of time that a particular unit is active during a sequence recall. In our network there are a two parameters that a priori are strong candidates to regulate the recall time: the transition weight $Exc_T$ by controlling how fast you activate the next unit in the sequence and the threshold $\theta$ by controlling how long it takes for the current to activate the next unit in a sequence. Here we show the exact results for $Exc_T$ but the reasoning for $\theta$ is analogous.

In order to calculate the recall time we make the following consideration. Once the unit $x_1$ becomes activated the filter $z_1$ starts increasing its value and the unit $x_2$ will be receiving an excitatory current $Exc_T \times z_1$. The point in time t at which the this current overcomes the threshold will be the recall time. Using that reasoning we can solve the following equation to calculate the recall time: $Exc_{T}(1 - e^{-\frac{t}{\tau_z}}) - \theta = 0$. If we do we end up with the following expression: $T_{recall} = \tau_z \log(\frac{Exc_T}{Exc_T - \theta})$. This expression tells us that we actually have a singularity at $Exc_T = \theta$, which for our purposes means that can make the recall time arbitrarily by setting this two parameters close enough. In our neural network however, there are more currents into play because of the all-to-all connectivity, we can include those corrections and we end up with an expression of the form $T_{recall} = \tau_z \log(\frac{Exc_T + Inh}{Exc_T - \theta})$ which also possess the after-mentioned property. We can see an illustration of the wide dynamical range of our system in figure \ref{Fig:dynamical_range} where we also show two examples of sequence recall with different $Exc_T$ values and therefore with different recall times.

\begin{figure}[h!]
\centering
\includegraphics[scale=0.25]{dynamical_range.eps}
\caption{The network posses a wide dynamical range. a) Here we compare the calculated thoretical expression with simulations with our network for different values of $Exc_T$. Note the singularity at $\theta$. b) The network activity with a short recall time. c) The network activity with a longer recall time ($Exc_T$ close to $\theta$).}\label{Fig:dynamical_range}
\end{figure}

\subsection{Learning Rule}

\begin{align}
\tau_w \dfrac{dw}{dt} &= (w_{max} - w) z_{pre}z_{post} + (w_{min} - w) (z_{post} - 1)z_{pre} 
\end{align}

\begin{figure}[h!]
\centering
\includegraphics[scale=1.3]{plasticity_diagram.eps}
\caption{This is how plasticity works}\label{Fig:plasticity diagram}
\end{figure}

\begin{figure}[h!]
\centering
\includegraphics[scale=0.3]{training_rule.eps}
\caption{The training in action}\label{Fig:epochs}
\end{figure}

\begin{figure}[h!]
\centering
\includegraphics[scale=0.3]{training_rule_quantities.eps}
\caption{The dynamics of the learning rule}\label{Fig:learning_quantities}
\end{figure}

\textbf{Training time}
\begin{itemize}
\item  $Exc_{self}$: The longer the training time the longer the filter is activated with itself and therefore it grows. But given that the pre and post filters do not coincide completely in the temporal domain there are negative contributions which prevent this from reaching the saturation value $w_{max}$
\item  $Inh$: In this case the the negative contributions come from the time that the units are not activated together with their counterparts which is most of the time (except in the brief transition period) and therefore the value saturates quickly to $w_{min}$.
\item $Exc_{T}$: In this case what matters is the transition time between two 
\end{itemize}

\subsection{Disambiguation}

\section{Discussion}

\section{Citation}
This ESANNV2.tex file defines how to insert references, both for
BiBTeX and non-BiBTeX users.  Please read the instructions in this
file.

>>>>

\documentclass[a4paper]{esannV2}
\usepackage[dvips]{graphicx}
\usepackage[latin1]{inputenc}
\usepackage{float}
\usepackage{subcaption}
\usepackage{amssymb,amsmath,array}
\graphicspath{{images/}}

@@

\begin{document}
%style file for ESANN manuscripts
\title{Sequence learning and recall in recurrent neural networks with memory traces}

%***********************************************************************
% AUTHORS INFORMATION AREA
%***********************************************************************
\author{Ram\'on H. Mart\'inez-Mayorquin $^1$, Anders Lansner $^1$ and Pawel Herman $^1$
%
% Optional short acknowledgment: remove next line if non-needed
% \thanks{This is an optional funding source acknowledgement.}
%
% DO NOT MODIFY THE FOLLOWING '\vspace' ARGUMENT
\vspace{.3cm}\\
%
% Addresses and institutions (remove "1- " in case of a single institution)
1- KTH Royal Institute of Technology \\
Department of Computational Science and Technology - Stockholm, Sweden.
%
% Remove the next three lines in case of a single institution
\vspace{.1cm}\\ 
}

@@

\begin{abstract} A rate-based recurrent network model equipped with a basic Hebbian learning rule including synaptic gating and memory traces is shown in this study to effectively learn sequences of activity patterns. The proposed computational framework facilitates understanding key mechanisms and the role of parameters in controlling the network performance and fundamental dynamics. We demonstrate among others how the sequence recall speed can be flexibly adjusted. 
\end{abstract}

\section{Introduction}
There is a growing interest in studying the capability to learn, store and flexibly recall sequential patterns in neural network models. Processing sequential information bears particular relevance to a wide scope of cognitive functionality and behavior in humans \cite{lashley1951problem}. Hebb proposed that sequential activation of neural cell assemblies, referred to as phases sequences, form the basis of the human thought process \cite{hebb2005organization}. The question of how a novel sequence of items is stored and recalled in the corrected order has been extensively studied in the context of memory systems among others. Besides intensive efforts in this direction undertaken in cognitive psychology \cite{hurlstone2014memory} the problem of sequence learning has also attracted a lot of theoretical interest. It has resulted in the development of different computational network models simulating the brain's capabilities to process sequence information. Among them are network approaches developed in the realm of rate based models that offer a generic context for sequence memory function \cite{amari1972learning}\cite{kohonen1977principle}\cite{sandberg2002bayesian}. The work presented here belongs to that category of studies and conceptually builds on \cite{sandberg2002bayesian}. In particular we rely on a simple recurrent neural network architecture and incorporate the concept of synaptic filtering into a basic Hebbian learning rule with synaptic gating \cite{andrew2003spiking} to study the network's capabilities to learn and recall sequences of memory items. The proposed rate based model constitutes a well-controlled framework for studying the fundamental synaptic mechanisms and identifying key parameters responsible for sequence processing capabilities in a dynamical connectionist system with all-to-all recurrent connectivity pattern. 
%
%A prominent share of this computaitonal endevaour is linked to spiking neural network approaches to learn precise patterns of spike sequences with biologically plausible local learning rules \cite{ans1994neural}\cite{dehaene1987neural}\cite{gutig2006tempotron}\cite{ponulak2010supervised}.

\section{Methods}

The recurrent neural network in Fig. \ref{Fig:diagrams} is composed of firing rate units with the input-output characteristics described by the logistic function $\Phi(x) = \frac{1}{1 + \exp(-Gx)}$ with gain G. The synaptic inputs are aggregated across the entire network through all-to-all connections weighted by synaptic weights $W$, and the activation of unit $U_i$, is then consistently determined by integrating synaptic inputs with the time constant $\tau_{m}$ (\ref{eq1}). To account for the temporal effect of synaptic smoothing we introduce the idea of z-filters  \cite{tully2016spike} that maintain the exponentially decaying traces of the activity with the time constant of $\tau_z$ (\ref{eq2}). This allows for keeping synaptic memory of the unit activity and influencing other post-synaptic units in the network even when the original pre-synaptic unit is not longer active.

@@

In this work we examine two types of learning rules, pre-synaptic (3) and post-synaptic (4)
gating \cite{andrew2003spiking}, which equip the proposed network with the capability to learn
how to recall sequential activity patterns.

\begin{align}
\tag{3 pre-synaptic}
\tau_w \dfrac{dw}{dt} &= (w_{max} - w) z_{pre}z_{post} + (w_{min} - w) (1 - z_{post})z_{pre} \label{eq:pre-synaptic rule}  \\
\tag{4 post-synaptic}
\tau_w \dfrac{dw}{dt} &= \underbrace{(w_{max} - w)}_{\text{saturation}} \underbrace{z_{pre}z_{post}}_{\text{Hebbian}} + \underbrace{(w_{min} - w)}_{\text{saturation}}\underbrace{ (1 - z_{pre})z_{post}}_{\text{synaptic-gating}} \label{eq:post-synaptic rule}
\end{align}

\begin{figure}[H]
\centering
\includegraphics[scale=0.8]{diagram_mixed.eps}
\caption{a) Scheme of the overall network connectivity of the system. Note that the inhibitory connections are only fully depicted for the first unit for the sake of clarity. b) Conceptual illustration of the mechanism of learning induced synaptic changes. Here the a black filling stands for an active unit. } \label{Fig:diagrams}
\end{figure}

The first term in (\ref{eq:pre-synaptic rule}) and (\ref{eq:post-synaptic rule}) is an associative Hebbian contribution \cite{hebb2005organization} and its role is to strengthen the connections between activated units. Importantly however, in comparison with the original formulation \cite{andrew2003spiking}, we use the z-filters instead of the unit activities. This provides the scope for making associative connections between units even when either of the two units is no longer active provided that its memory trace is retainedby the synapse. As mentioned earlier, the synaptic trace decays exponentially with the time constant of $\tau_z$ (\ref{eq2}), and to preserve serial recall order we introduce asymmetrical time constants for z-pre and z-post : $\tau_{z_{pre}} > \tau_{z_{post}}$. The second term is called synaptic gating \cite{andrew2003spiking}, which facilitates the formation of inhibitory connections whenever the activity (or its z-filtered trace) of one unit is not accompanied with the activity of the other unit. We examine two variations of this term to explore the functional implications of both pre- and post-synaptic gating. Finally, the weights are kept between $w_{min}$ and $w_{max}$ limits. The nature of synaptic changes under the synaptic gating rule is illustrated in Fig. \ref{Fig:diagrams}.

%\begin{figure}[h!]
%\centering
%\includegraphics[scale=1.3]{plasticity_diagram.eps}
%\caption{Conceptual illustration of the mechanism of learning induced synaptic changes. Here the a black filling stands for an active unit a) Hebbian associative plasticity that implies a weight increase when both pre- and pos-synaptic units are simultaneously active b) In the case of the post-synaptic rule the weight is decreased if the post-synaptic unit itself active without the pre-synaptic unit being active. c) In the case of the pre-synaptic rule the weight is decreased if the pre-synaptic unit is itself active without the post-synaptic unit being active. d) If neither pre- nor post-synaptic units are active, there is no weight modification.}\label{Fig:plasticity diagram}
%\end{figure}
%
As a result of learning described above, three types of connections emerge from each unit (Fig. \ref{Fig:diagrams}): two excitatory connections - self-reinforcing unit activity, $Exc_{self}$, and exciting the following unit in the sequence (contributing to the sequential structure of the activity) $Exc_T$, and a range of inhibitory connections, $Inh$, to all the remaining units (among inhibitory connections of special relevance are the backwards inhibitory projections that suppress the unit activity once the next unit in the sequence gets activated).

@@

To recall a particular sequence we cue the first sequence item (unit) by injecting a short input lasting around $\tau_z$. After this initial cue, if the weights have the right configuration as a result of training, the neural network continues to visit subsequent items in the learned sequence to the very
end, as shown in Fig. \ref{Fig:recall}.

\begin{figure}[h!]
\centering
\includegraphics[scale=0.20]{recall.eps}
\caption{Demostration of the successful sequence recall with the corresponding connectivity matrix. a) The activity of the units over time. b) The currents that each unit receives as the sequence progresses. c) the z-filters activations for each unit. d) The connectivity matrix $W$. The weights of the matrix $W$ can be divided into the excitatory connections: the ones on the diagonal that stabilize the unit $Exc_{self}$, the ones under the diagonal that induce transitions $Exc_{T}$ and finally the ones that stabilize the unit  and the inhibitory connections $Inh$ which dominate the matrix $W$. We use different colors to represent different items. }\label{Fig:recall}
\end{figure}

The dynamics of the sequential activity propagation can be easily explained. After the first unit $U_1$ has been activated by the external input, the corresponding filter $z_1$ starts following the activity $x_i$. In consequence, the subsequent unit $U_2$ receives a growing current with value $z_1 Exc_T$ because of the forward connectivity. This process continues until the point in time when $Exc_{T}(1 - e^{-\frac{t}{\tau_z}}) - \theta = 0$ implying that $U_2$ receives enough current to become activated. This in turn prompts the filter $z_2$ to follow the activity $x_2$ initializing again the process described above. On the other hand as soon as the synaptic trace corresponding to $x_2$ starts growing, the preceding unit $U_1$ beings to receive an inhibitory current $z_2 Inh$ due to the backward inhibitory projection. This in turn suppresses the activity of unit $U_1$ facilitating the propagation of the sequence. Finally once the unit $U_1$ is completely suppressed the excitatory current mediated by $Exc_T$ is diminished, which in most cases would also suppress the activity $x_2$ if it was not for the self-sustaining current $z_2 Exc_{self}$. In short, the term $Exc_T$ ensures the transition of the activity from one network state (sequence item) to another one (the next item in the sequence) whereas the competition between $Exc_{self}$ ad $Inh$ first lock the pattern unit and then finally suppresses it enabling the sequence to continue its trajectory.

With the knowledge above in mind we can envision the following  failure scenarios for the sequential dynamics. First, if the magnitude of $Exc_{self}$ weight is small the self-excitatory current is not sufficiently strong to overcome the threshold and the sequence stalls. Second, if $Exc_{self}$ is large enough to keep the unit active but the $Exc_{T}$ weight is weak, the next unit in the sequence does not become activated and the sequence remains stuck forever in the same state corresponding to the first pattern. Finally if both $Exc_{self}$ and $Exc_T$ are strong enough to sustain and propagate the activity respectively but the inhibitory weight $Inh$ is not sufficiently large to suppress the activity, all the units in the sequence become co-activated at the end.

@@

A common problem in sequence learning is to scale the time duration of the sequence according to the contextual demands \cite{murray2017learning}. For a neural network it means that the dynamics of sequence recall can be flexibly controlled with the network parameters after the training process is completed. The recall time can be defined as the temporal length of the activation of the individual units during the sequence retrieval. In the network there are two key parameters that are directly implicated in the regulation of the recall time: the transition weight, $Exc_T$ and the threshold, $\theta$, which control how long it takes to recruit the next unit in a sequence.  By taking into account the competition between all the currents involved and the threshold we can calculate exactly the point at which the unit will become suppressed:

\begin{align}
T_{recall} = \tau_z \log\Big(\frac{Exc_T + Inh}{Exc_T - \theta}\Big) \label{eq:recall_complex}
\end{align}

This expression has a singularity at $Exc_T = \theta$, which implies that the recall time can be made arbitrarily long by setting $Exc_T$ close to the threshold. Fig. \ref{Fig:dynamical_range} illustrates the wide dynamical range of the recall time in our network model.

\begin{figure}[h!]
\centering
\includegraphics[scale=0.20]{dynamical_range.eps}
\caption{The network posses a wide dynamical range. a) Here we compare the calculated thoretical expression with simulations with our network for different values of $Exc_T$. Note the singularity at $\theta$. b) The network activity with a short recall time. c) The network activity with a longer recall time ($Exc_T$ close to $\theta$).}\label{Fig:dynamical_range}
\end{figure}

\subsection{Learning Rule}
Our learning protocol consists in fixing (clamping) each unit of the sequence in succession for a given amount of time (training time). We perform this process an epochs number of times. It is important to note that while learning there is a period of one second between each epoch or presentation that serves the purposes of avoiding self-interference.

We illustrate the results of training in Fig. \ref{Fig:epochs} with two connectivity matrix capable of recalling, note the effects of $\tau_z$ in the strength and extent of the $Exc_T$ effects. Furthermore we can appreciate that the weights converge and stabilized with few epochs of training.

\begin{figure}[H]
\centering
\includegraphics[scale=0.25]{training_rule.eps}
\caption{The effect of training with the learning rule. a) An example of training with a small value of $tau_z=0.050s$. b) Same procedure as in a but with a longer time constant $\tau_z=0.150s$. c) evolution in time of the connection weights from the first to the second unit $Exc_T$, the one from the second back to the first unit $Inh$ and finally of the second unit into itself $Exc_{self}$. Five epochs of length $2s$ are shown here, note that this is enough time for the weights to stabilize.}\label{Fig:epochs}
\end{figure}

%\begin{figure}[H]
%\centering
%\includegraphics[scale=0.3]{training_rule_quantities_pre_rule_True.eps}
%\caption{Learning dynamics of the pre-synaptic learning rule.}\label{Fig:learning_quantities_pre}
%\end{figure}
%
%
%\textbf{Training time}
%\begin{itemize}
%\item  $Exc_{self}$: The longer the training time the longer the filter is activated with itself and therefore it grows. But given that the pre and post filters do not coincide completely in the temporal domain there are negative contributions which prevent this from reaching the saturation value $w_{max}$
%\item  $Inh$: In this case the the negative contributions come from the time that the units are not activated together with their counterparts which is most of the time (except in the brief transition period) and therefore the value saturates quickly to $w_{min}$.
%\item $Exc_{T}$: In this case what matters is the transition time between two 
%\end{itemize} 
%
%
%\begin{figure}[H]
%\centering
%\includegraphics[scale=0.3]{training_rule_quantities_pre_rule_False.eps}
%\caption{Learning dynamics of the post-synaptic learning rule}\label{Fig:learning_quantities_post}
%\end{figure}

%\section{Discussion}
